---
title: "Lab 1"
author: "Nate Beebe (Worked with Matthew Raitano and Zack Barnes)"
date: "Due on 02/07 at 11:59 PM"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


**Instructions:** This lab report needs to be professional. Only report relevant and finalized code. Your writing should be concise and void of spelling errors. Use code chunk options to hide unnecessary messages/warnings. Your report should be reproducible. Reports that involve simulations need to have the random seed specified so that simulation results are reproducible. You are allowed to work on this lab assignment in groups of 2-3. You still need to submit an individual lab report if you do work in a group, and you need to list your collaborators. \vspace*{0.5cm}


**Question 1** In lecture it was demonstrated that baseball is a game of offense, pitching, and defense with a regression model that considered expected run differential as a function of explanatory variables OPS, WHIP, and FP. Do the following:

 - Fit a similar regression model with runs as the response variable. Report problems with this model. Investigate problematic residuals to discover what went wrong. Fix the problem with this model by adding categorical variable(s) to the list of explanatory variables. Briefly explain what went wrong. 
 
```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(Lahman)
library(tidyverse)
library(broom)
#Creating dataset
dat = Teams %>% 
	select(yearID, franchID, W, L, G, AB, H, X2B, X3B, HR, BB, HBP, SF, 
				 HA, HRA, BBA, SOA, IPouts, FP, R, RA) %>% 
	filter(yearID >= 1900) %>%
	replace_na(list(HBP = 0, SF = 0)) %>% 
	mutate(RD = (R - RA) / G) %>% 
	mutate(OBP = (H + BB + HBP)/(AB + BB + HBP + SF)) %>%  
	mutate(SLG = (H + X2B + 2*X3B + 3*HR)/AB) %>% 
	mutate(OPS = OBP + SLG) %>% 
	mutate(WHIP = 3*(HA + BBA)/IPouts)

#Fitting model
model1 = lm(R ~ OPS + WHIP + FP, dat)

summary(model1)
plot(model1)

#Adding residuals to check
dat2 = dat %>% 
  mutate(resid = model1$residuals) %>% 
  arrange(resid)
dat2[1:31,c(1,27)]

#Adding COVID T/F Variable
dat3 = dat %>% 
  mutate(COVID = yearID == 2020)

#Fitting the new model
model2 = lm(R ~ OPS + WHIP + FP + COVID, dat3)
summary(model2)

```
The shortened 60 game 2020 season due to COVID-19 did not fit the mold of the model that used seasons with more than 100 games for every other year. The total number of runs was significantly lower in this year than any other. The residuals for each of the 30 teams during that year were the 30 highest absolute value residuals in the entire model (and over 30 units greater than the next closest). Adding the 'COVID' variable allows for a big change in the model when the year was 2020.
 
 
 - We can significantly improve the regression model in the notes through a principled rescaling of OPS, WHIP, and FP. Split the Teams data frame by \texttt{yearID} and, for each year, create variables \texttt{OPSscale = OPS/avgOPS}, \texttt{WHIPscale = avgWHIP/WHIP}, and \texttt{FPscale = FP/avgFP} which require you to first create league average variables \texttt{avgOPS}, \texttt{avgWHIP}, and \texttt{avgFP}. Fit the linear regression model with runs differential as the response and explanatory variables \texttt{OPSscale}, \texttt{WHIPscale}, and \texttt{FPscale}, and report relevant output. Why does this model perform so much better than the model in the notes? Support your answer. 

```{r, message=FALSE, warning=FALSE, echo=FALSE}
#Creating scaled dataset
dat4 = dat %>% 
  mutate(X1B = H - X2B - X3B - HR,
         IP = round(IPouts/3 ,2))
scale = dat4 %>% 
  group_by(yearID) %>%
  mutate(avgOBP = round((sum(H) + sum(BB) + sum(HBP))/ (sum(AB) + sum(HBP) + sum(BB) + sum(SF)),3),
         avgSLG = round((sum(X1B)+2*sum(X2B)+3*sum(X3B)+4*sum(HR))/sum(AB), 3),
         avgOPS = avgSLG + avgOBP,
         avgWHIP = round((sum(BB)+sum(H))/sum(IP),3),
         avgFP = round(sum(FP)/n(),3)) %>% 
  mutate(OPSscale = OPS/avgOPS,
         WHIPscale = avgWHIP/WHIP,
         FPscale = FP/avgFP)

#Fitting the new model
model3 = lm(RD ~ OPSscale + WHIPscale + FPscale, scale)
summary(model3)
```
This model works better because it compares each team relative to their peers. Baseball's offensive output has changed a lot over its lifespan, and as it evolved, so did offensive and defensive output (leading to differences in run differential). A team from 1900 shouldn't be put on the same scale as a team from 2023.


**Question 2** Choose 3 batters and 3 pitchers that have played in at least 10 seasons and do the following: 

Batters selected: David Ortiz, Salvador Perez, and Lorenzo Cain;
Pitchers selected: Madison Bumgarner, Wade Davis, and Jacob deGrom.

 - Display the seasonal statistics for these players. The following statistics should be included for batters (derivations of unconventional statistics are in parentheses): year, G, AB, R, H, X2B, X3B, HR, RBI, SB, CS, SBpct (SB / (SB + CS)), BB, SO, OBP, SLG, OPS. The following statistics should be included for pitchers: year, W,  L, IPouts, H, ER, HR, BB, HBP, SO, ERA, WHIP, SOper9 (SO / IP * 9), SOperBB (SO / BB). These statistics can be found in or computed from statistics that are found in the \texttt{Batting} and \texttt{Pitching} dataframes in the \texttt{Lahman} package.
 
```{r, message=FALSE, warning=FALSE, echo=FALSE}
#Matching players to statistics
Dataset = People %>% 
  select(playerID, nameFirst, nameLast) %>% 
  left_join(Batting) %>% 
  mutate(year = yearID) %>% 
  select(nameFirst, nameLast, year, G, AB, R, H, X2B, X3B, HR, RBI, SF, SB, CS, BB, HBP, SO)

#Creating rate statistics
YearlyDataset = Dataset %>% 
  mutate(OBP = round((H + BB + HBP) / (AB + BB + HBP + SF), 3),
         X1B = (H - X2B -X3B - HR)) %>% 
  mutate(SLG = round((X1B + X2B * 2 + X3B * 3 + HR * 4)/AB, 3)) %>% 
  mutate(OPS = SLG + OBP) %>% 
  mutate(SBpct = round(SB/(CS+SB), 3)) %>% 
  select(nameFirst, nameLast, year, G, AB, R, H, X2B, X3B, HR, RBI, SB, CS, SBpct, BB, SO, OBP, SLG, OPS)

#Selecting the players I want
Batters = YearlyDataset %>% 
  filter(nameFirst == "Lorenzo" & nameLast == "Cain" | nameFirst == "Salvador" & nameLast == "Perez" | nameFirst == "David" & nameLast == "Ortiz")

Batters

#Repeating process for pitchers
PitcherSet = People %>% 
  select(playerID, nameFirst, nameLast) %>% 
  left_join(Pitching) %>%
  mutate(year = yearID) %>% 
  select(nameFirst, nameLast, year, W, L, IPouts, H, ER, HR, BB, HBP, SO, ERA,
         )

PitcherYearly = PitcherSet %>% 
  mutate(IP = round(IPouts / 3, 2)) %>% 
  mutate(WHIP = round((BB + H)/ IP, 3)) %>% 
  mutate(SOper9 = round(SO / IP * 9, 3)) %>% 
  mutate(SOperBB = round(SO / BB, 3)) %>% 
  select(nameFirst, nameLast, year, W, L, IPouts, H, ER, HR, BB, HBP, SO, ERA, WHIP, SOper9, SOperBB)
  
  
Pitchers = PitcherYearly %>% 
  filter(nameFirst == "Madison" & nameLast == "Bumgarner" | nameFirst == "Wade" & nameLast == "Davis" | nameFirst == "Jacob" & nameLast == "deGrom")

Pitchers

  
```
 

 - Create career stat lines for each of the players that you selected. Be careful about how these statistics are calculated.
 
```{r, message=FALSE, warning=FALSE, echo=FALSE}
#Summarizing statistics for career
Dataset %>%
  filter(nameLast == "Ortiz" & nameFirst == "David" | nameLast == "Perez" & nameFirst =="Salvador" | nameLast == "Cain" & nameFirst == "Lorenzo") %>% 
  group_by(nameLast) %>% 
  summarise(across(G:SO, sum)) %>%
    mutate(OBP = round((H + BB + HBP) / (AB + BB + HBP + SF), 3),
         X1B = (H - X2B -X3B - HR)) %>% 
  mutate(SLG = round((X1B + X2B * 2 + X3B * 3 + HR * 4)/AB, 3)) %>% 
  mutate(OPS = SLG + OBP) %>% 
  mutate(SBpct = round(SB/CS, 3))


PitcherSet %>% 
  filter(nameFirst == "Madison" & nameLast == "Bumgarner" | nameFirst == "Wade" & nameLast == "Davis" | nameFirst == "Jacob" & nameLast == "deGrom") %>% 
  group_by(nameLast) %>% 
  summarise((across(W:SO, sum))) %>% 
  mutate(IP = round(IPouts / 3, 2)) %>% 
  mutate(WHIP = round((BB + H)/ IP, 3)) %>% 
  mutate(SOper9 = round(SO / IP * 9, 3)) %>% 
  mutate(SOperBB = round(SO / BB, 3)) %>%
  mutate(ERA = round((9 * ER/ IP), 2)) %>% 
  select(nameLast, W, L, IP, H, ER, HR, BB, HBP, SO, ERA)
  

  
```
 
 
 - Provide a plot for career trajectories for one batting and one pitching statistic of your choice. These are two separate graphics, one for the batters and one for the pitchers. The graphics that you produce should display the trajectories of the 3 batters and the 3 pitchers. Provide interesting commentary on your graphic.
 
```{r, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(data = Batters, aes(x = year, y = OPS, color = nameLast)) + 
  geom_line()+
  labs(
    x = "Season",
    y = "OPS",
    color = "Player"
  )
```
David Ortiz started his career by far the earliest of these three players, and he had the best OPS of the three for the entire time they overlapped. He is known as a power hitter, and having the highest OPS of the three reflects that. Lorenzo Cain was far from a power hitter, but he did finish third in AL MVP voting in 2015 (where his OPS was the highest of his career). Perez is considered a power hitter, but he has had some issues reaching base in his career which is reflected in OPS. In 2020 he was honored with All-MLB First-Team honors at catcher, and in 2021 he hit 48 home runs. Those two seasons are the peak and the next highest point in his OPS graph, respectively.


```{r, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(data = Pitchers, aes(x = year, y = WHIP, color = nameLast))+
  geom_line()+
  labs(
    x = "Season",
    y = "WHIP",
    color = "Player"
  )
```
Wade Davis had a stretch from 2013-2015 where he was arguably the best relief pitcher the game had ever seen. That is reflected here where his WHIP is the lowest of his career. Madison Bumgarner had a lengthy career and he was known to be a much better pitcher in the playoffs (playoff data is not included in this chart). His WHIP got noticably worse as he reached the tail end of his career. Jacob deGrom is known as one of the best pitchers in baseball when he is healthy. His WHIP from 2018 onward does not go above 1.00 which is incredibly impressive. 
 


**Question 3** Exercise 1.2 in the online version of Analyzing Baseball Data with R. Exercise 2 on page 27 of the textbook.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
Gibson = Pitching %>% 
  filter(playerID == "gibsobo01" & yearID == 1968) %>% 
  mutate(IP = round(IPouts / 3, 2))

Gibson %>% 
  mutate(CompletedRatio = CG / GS) %>% 
  select(GS, CG, CompletedRatio)

Gibson %>% 
  mutate(KBB = round(SO / BB, 4)) %>% 
  select(SO, BB, KBB)

Gibson %>%
  select(IP)

Gibson %>% 
  mutate(WHIP = round((BB + H)/IP, 3)) %>% 
  select(WHIP)

```


a. 28/34 games were completed by Gibson (0.8235).
b. Gibson's strikeout to walk ratio was 268/62 or 4.3226.
c. He pitched 304 and 2/3 innings.
d. Gibson had a WHIP of 0.853.

 
**Question 4** Exercise 1.3 in the online version of Analyzing Baseball Data with R. Exercise 3 on page 27 of the textbook.

a. The game was 2 hours and 19 minutes long.
b. Because it is the second game of a doubleheader and the attendance was logged in the first game.
c. They had 3 extra-base hits.
d. It was 0.375 (12/32).

